Finding and controling biases in dark matter halo concentrations estimates

### 2016-07-04 paper/ms.tex
new references , new order

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title{A less biased dark matter halo concentration using the integrated mass profile} 
\author{
  C. N. Poveda-Ruiz \thanks{cn.poveda542@uniandes.edu.co}$^{1}$
  J. E. Forero-Romero \thanks{je.forero@uniandes.edu.co}$^{1}$
  J. C. Mu\~noz-Cuartas \thanks{juan.munozc@udea.edu.co}$^{2}$
}

@@

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.   
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of the new method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
We design another simple test that can be used on any
data-set to estimate and correct the biases when estimating the
concentration at low particle numbers. 
That test shows that the new method is robust with respect to changes
in halo particle number. 
However, the velocity method over-predicts concentration values by a
factor of $25\%$ values for low particle numbers.
Finally, we also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{10}\hMsun < M_{h} <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
There is a small deviation at low masses, $M<10^{12}\hMsun$, where
the new method yields lower median concentration values by $25\%$
compared to the velocity methods; accounting for the bias in the
velocity method brings the results into agreement.  
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

The internal DM distribution in a halo is usually parameterized
through the density profile.  In a first approximation this profile is
spherically symmetric; the density only depends on the radial
coordinate.  One {\bf of} the most popular radial parameterizations is
the Navarro-Frenk-White (NFW) profile \citep{NFW}.  If one is not
interested in the very central region of the dark matter halo (where
galaxy formation takes place, and where the effects of baryon physics
on dark matter distribution are still unknown) one can also assume
that this profile is universal.  \citep{Navarro2010}.  This profile is
a double power law in radius, where the transition break happens at
the so-called scale radius, $r_s$.  The ratio between the scale radius
and the halo virial radius $R_v$ is known as the concentration
$c=R_v/r_s$.

@@

The first method is straightforward to apply but presents two
disadvantages.  First, it requires a large number of particles in
order to have a proper density estimate in each bin.  This makes the
method robust only for halos with at least $10^3$ particles.  The
second problem is that there is not a way to estimate the optimal
radial bin size, different choices may produce different results for
the concentration.

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.  This approach has two advantages with
respect to the above mentioned methods.  It does not involve any data
binning and does not throw away data points.  This translates into a
robust estimate even at low resolution/particle numbers.  Furthermore,
since the method does not require any binning, there is no need to
tune numerical parameters.  This method provides a new independent
method to estimate the concentration parameter.  Here we provide a
detailed description of the method and a comparison against the two
traditional techniques by measuring its performance both on mock and
N-body halos.

@@

We use data from the Bolshoi cosmological simulation that follows the
non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.  The
data is publicly available at \url{http://www.cosmosim.org/}.  More
details about the structure of the database and the simulation can be
found in \citep{2013AN....334..691R}.

@@

\subsubsection{Convergence Test using Downsampling}

We perform first a convergence test on the N-body data to find
possible biases in the concentration estimates as a function of
particle number.  
We select $8$ halos from the simulation that have at
least $3\times 10^5$ particles.  
We then randomly downsample the total number of particles in each halo
by factors of $10, 20, 45, 100, 215, 460, 1000$ to measure its
concentration with both the integrated mass and the velocity
method. For each downsampling factor we repeat the operation $100$
times.

@@

@@

\bibliographystyle{mn2e}
\bibliography{references}

>>>>

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title{Finding and controling biases in dark matter halo concentrations estimates}
\author{
  C. N. Poveda-Ruiz \thanks{cn.poveda542@uniandes.edu.co}$^{1}$
  J. E. Forero-Romero \thanks{je.forero@uniandes.edu.co}$^{1}$
  J. C. Mu\~noz-Cuartas \thanks{juan.munozc@udea.edu.co}$^{2}$
}

@@

\begin{abstract}
We use bootstrapping to estimate the bias of concentration
estimates on N-body dark matter halos as a function of particle
number. 
We find that algorithms based on the maximum radial velocity
and radial particle binning have increaing biases with decreasing 
particle numbers.
For instance, these methods overestimate concentrations by 
$15\%-20\%$ for halos sampled with $200$ particles.  
To overcome this bias we propose a new algorithm based on the
integrated mass profile. 
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.
This method reduces the bias to $2\%-3\%$ for halos sampled with
$200$ particles.  
To keep the biases of the velocity and density methods to the same
low levels as in the new method, dark matter halos should be sampled
with at least $\sim 2000$ particles. 
We also show that the mass-concentration relationship could be
shallower than expected from numerical experiments once the biases 
of the different concentration measurements are taken into account.
From these results we believe that bootstrapping and the
concentration estimation based on the integrated mass are
valuable tools to probe the internal structure of dark matter halos
in numerical simulations.
\end{abstract}

@@

The internal DM distribution in a halo is usually parameterized
through the density profile.  In a first approximation this profile is
spherically symmetric; the density only depends on the radial
coordinate.  One of the most popular radial parameterizations is
the Navarro-Frenk-White (NFW) profile \citep{NFW}.  
This profile can be considered as universal \citep{Navarro2010}, assuming that 
one is not interested in the very central region of the dark matter halo where
galaxy formation takes place, and where the effects of baryon physics
on dark matter distribution are still unknown.
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  The ratio between
the scale radius and the halo virial radius $R_v$ is known as the
concentration $c=R_v/r_s$.

@@

The first method is straightforward to apply but presents two
disadvantages.  First, it requires a large number of particles in
order to have a proper density estimate in each bin.  This makes the
method robust only for halos with at least $10^2$ particles.  The
second problem is that there is not a way to estimate the optimal
radial bin size, different choices may produce different results for
the concentration.

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.  
This approach has two advantages with respect to the above mentioned methods.  
It does not involve any data binning and does not throw away data points.  
This translates into a robust estimate even at low resolution/particle numbers.  
Furthermore, since the method does not require any binning, there is no need to
tune numerical parameters.   
This method provides a new independent method to estimate the
concentration parameter.

In order to asses the quality of the three different methods we use
bootstraping on halos that are sampled with $10^{5}$ particles.
This allows us to estimate the bias and standard deviation on the
concentration estimates as a function of particle number.

@@

We use data from the Bolshoi cosmological simulation that follows the
non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.  
The cosmological parameters use a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The
data is publicly available at \url{http://www.cosmosim.org/}.  More
details about the structure of the database and the simulation can be
found in \citep{2013AN....334..691R}.

@@

We also use public data from the Via Lactea simulation project
\citep{2008Natur.454..735D}.  
This simulation was run for a single isolated halo with a virial mass
of the order of $10^{12}$\hMsun using the parallel tree code PKDGRAV
\citep{2001PhDT........21S}.  
The simulation used $\sim 2\times 10^{8}$ particles to resolve this
region.
The cosmological parameters are diffferent from those in the Bolshoi
simulation, with a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.238$ and a normalization of the power spectrum
$\sigma_8=0.74$. 
The data available to the public corresponds to a downsampled set 
of $10^5$ particles which corresponds to a particle mass of
$2.24\times 10^{7}$\hMsun.

We use a set of two very different simulations to demonstrate that the 
results of our bootstraping tests are independent of the kind of
parent simulation.

\section{Bootstrapping to estimate biases}
\label{sec:bootstrapping}

Avoiding biases with varying simulation resolution is one of the most
important goals of a robust method to estimate halo concentration. 
We propose bootstraping individual high resolution halos to quantify these
possible biases.
In this experiment we take halos sampled with at least $10^{5}$
particles and subsample each halo by factors of $2$ up to $10^{3}$ to
measure the concentration every time the halo is resampled.

@@

We select $8$ halos from the simulation that have at
least $3\times 10^5$ particles.  
We then randomly downsample the total number of particles in each halo
by factors of $10, 20, 45, 100, 215, 460, 1000$ to measure its
concentration with both the integrated mass and the velocity
method. For each downsampling factor we repeat the operation $100$
times.

@@

\bibliographystyle{apj}
\bibliography{references}
